\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{blindtext}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{booktabs}
\usepackage{float}
\usepackage{bbm}
\graphicspath{ {./images/} }
\usepackage[bottom=0.5cm, right=1.5cm, left=1.5cm, top=1.5cm]{geometry}

\newtheorem{theorem}{Theorem}
\newtheorem{exercise}{Exercise}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}[section]

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\newcommand{\p}{\ensuremath{\mathbb{P}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\1}{\ensuremath{\mathbbm{1}}}
\newcommand{\B}{\ensuremath{\mathbbm{B}}}

\title{Lecture 1}
\author{Gidon Rosalki}
\date{2025-03-23}


\begin{document}
\maketitle
\section{Introduction}\label{sec:Introduction} % (fold)
ENIAC - the first computer. Used vacuum tubes, which are gates that operate
based off electronic emissions of heated wires. Was about 27 tonnes, 160KW of
power, used 18,000 vacuums tubes which needed to be changed at a rate of about
2,000 a day. It was the first computer to find the first 2,000 digits of pi.
"There is no reason anyone would want a computer in their home" - Ken Olsen \\
Computers consist of IO devices, storage devices which may be volatile (do not
survive power disconnection eg RAM), and non volatile (storage, HDDs, SSDs,
etc.), the processing unit CPU, the operating system, device drivers, and
application software. We will cover the CPU, and application SW, and mention
storage devices, and the OS. \\
Moore's Law (1965): Intel co-founder Gordon Moore predicted in 1965 that the
number of transistors on a chip will double about every 2 years, which has been
the case for 40 years now. This was achieved by telling the engineers that they
may go home once they have achieved a doubling in 2 years. If they achieved
more, then it was spread out to a doubling every 2 years. \\
The AAA paradigm: Every computer scientist should master Architecture,
Algorithms, and Applications. \\
% section Introduction (end)

\section{Semiconductors}\label{sec:Semiconductors} % (fold)
The first transistors were made in 1947 by John Bardeen and Walter Brattain in
Bell Labs, and was made of germanium. A transistor is a solid state switch,
which may be switched through the application of electric charge. In 1958 Jack S
Kilby at Texas Instruments developed the first integrated circuit (again using
germanium). It contained 5 components, of 3 different types: Transistors,
resistors, and capacitors. \\
The periodic table was created first by Mendeleev in 1869. It was a way to
organise the chemical elements by weight, but these days is organised by
electron configuration. Each row of the table is a 'layer' of electrons, and for
the most part the elements in the same column have the same number of electrons
in their outer shell. Bohr suggested the Bohr model, where the nucleus contains
the protons and neutrons, with electrons in shells around it, where the first
shell has up to 2, second and third up to 8, and the fourth and fifth up to 18,
and so on. \\
Atoms "want" a full outer shell, and this is achieved through forming bonds.
These can be covalent bonds, where an electron is shared between the atoms,
and ionic bonds, which bonds charged ions together. (There is also metallic
bonding not discussed in this course). \\
We consider that electrical current is the flow of electrons (or holes). The
electrons are viewed by convention as moving from positive to negative. This
is analogous to water flow. Voltage is the potential difference. The units are
volts, and we use the term "ground" to indicate the point of lowest potential.
This is also analogous to water potential energy (consider a water tower instead
of a battery). The charge is the amount of electrons that are stored, and
capacitance is the ability to store charge. When a charge is stored in a
capacitor, a voltage is applied on it. A capacitor is analogous to a water
tower. In an electrical circuit current will flow from high potential to lower
potential. Resistance will impact the rate of flow, higher resistance slows the
flow, where lower resistance allows for greater flow. Electrical circuits may
also include branches. Flow is either from a power supply, of from a capacitor to
ground. \\
We will use capacitors in order to hold $0$ or $1$. A $1$ is represented by a
capacitor with high voltage, and $0$ is low voltage. Logical values will be
represented with 0s and 1s in binary, and we will generally use hexadecimal to
represent groups in a short format. \\
We use silicon to create our chips. Pure silicon is a semiconductor and does not
conduct electrical current well because it has few free charge carriers. Silicon
dioxide ($SiO_2$, or sand) is an insulator, and conducts no electrical current.
We use doping, an intentional introduction of impurities into a semiconductor to
modulate its properties. We create P-type, and N-type. P-type has extra
aluminium in the semiconductor, which reduced the number of available electrons,
meaning that there are extra holes. N-type has added phosphorous, which is an
electron donor, and increases the number of free electron in the substance. \\
If we connect together p-type and n-type, we can create a PN junction, which
conducts in one direction, and blocks current in the other direction, which we
call a \textbf{diode}. % TODO direction of blocking and allowing
% section Semiconductors (end)

\section{MOS-FET transistor structure and operation}\label{sec:MOS-FET transistor structure and operation} % (fold)
MOS-FET = Metal Oxide Semiconductor - Field Effect Transistor. \\
% TODO add some images from the presentation
Contains the terminals Source, Drain, and Gate. They are constructed from an
insulator, and two NP junctions, and a capacitor which is the gate body. There
are two variates, N-Channel (NMOS), and P-Channel (PMOS). \\
When considering transistors, there are a few parts to consider. L is the length
of the channel / gap between the source and gate. W is the width of the
transistor (along the source/gate), and Lox is the oxide thinness, which impacts
the gate capacitance. \\
NMOS is sued as a voltage controlled switch. The gate controls whether current
can flow between the source and the drain. When the gate reads $0$, we have two
charged NP junctions, and thus current cannot flow between S and D. When the
gate reads $1$, then there is an N like channel between S and D, and the current
can flow. PMOS operates in the opposite direction.
% section MOS-FET transistor structure and operation (end)

\section{Manufacture}\label{sec:Manufacture} % (fold)
Making wafers is done by melting pure silicon, and adding a very small amount of
the dopant. You then insert and rotate an ingot, a starting crystal on which the
doped silicon crystallises. Once this process is completed, the crystals are cut
and polished into wafers. To process the wafers, geometries are defined using
light, and a film coating that is light sensitive. Generally uses a wavelength
of $<1nm$. Each step creates a mask, does a chemical/physical step on the light
area (etching, doping, depositing), clean the mask, and then flattened.
% section Manufacture (end)

\end{document}
